---
title: "Cleaning"
format: html
---

## Goals of this notebook

Our data all comes from the [MLS Players Association](https://mlsplayers.org/resources/salary-guide) but it was published in different formats throughout the years.

- We are starting with some historical data already cleaned by Jeremy Singer-Vine for the Data is Plural newsletter: [Github archive](https://github.com/data-is-plural/mls-salaries). It goes from 2007 to 2017.
  + It's interesting for 2017 the PDF DIP has is different than what I downloaded in that DIP has the pool MLS players but mine does not.
- Then our 2018 to 2021 data is converted from PDF using [Tablula](https://tabula.technology/) into CSVs, but this data will need headers, etc.
- Then out 2022 and 2023 data comes as Excel files.

Once all the data is imported, we'll still need to do some normalization: - All columns will need the same datatype, etc. - We'll want to normalize team names. Sometimes team initials are used, and other times full names are used. I'd like to have both. I'll probably build something in Google Sheets. - I'd like to add conference affiliations. I'll likely have to build this and might only do for recent years.

## Setup

```{r}
#| label: setup
#| warning: false
library(tidyverse)
library(readxl)
library(janitor)
```

## Import the archive data

This brings in the Data is Plural files, which should all be the same.

```{r}
#| label: dip-import

dip_data <- read_csv(c(
  "data-processed/data-is-plural/mls-salaries-2007.csv",
  "data-processed/data-is-plural/mls-salaries-2008.csv",
  "data-processed/data-is-plural/mls-salaries-2009.csv",
  "data-processed/data-is-plural/mls-salaries-2010.csv",
  "data-processed/data-is-plural/mls-salaries-2011.csv",
  "data-processed/data-is-plural/mls-salaries-2012.csv",
  "data-processed/data-is-plural/mls-salaries-2013.csv",
  "data-processed/data-is-plural/mls-salaries-2014.csv",
  "data-processed/data-is-plural/mls-salaries-2015.csv",
  "data-processed/data-is-plural/mls-salaries-2016.csv",
  "data-processed/data-is-plural/mls-salaries-2017.csv"),
  id = "file_name"
) |> 
  mutate(
    year = str_sub(file_name, 44, 47)
  ) |> 
  relocate(year) |> # moving year to front
  select(!file_name) |>  # dropping file_name
  rename(club_short = club)

dip_data |> head()
```

### Cleaning clubs

There are times when club abbreviations vary. We will fix those here.

Checking them:

```{r}
dip_data |> 
  count(club_short)
```

Looking through this, I want to make a number of changes.

- TOR changed to TFC
- NY to NYRB since that's what is used after NYCFC arrived
- None changed to NA
- Pool changed to Major League Soccer
- POOL changed to Major League Soccer

This makes the changes.

```{r}
dip_changed <- dip_data |> 
  mutate(
    club_short_clean = recode(
      club_short,
      "TOR" = "TFC",
      "NY" = "NYRB",
      "None" = NULL,
      "POOL" = "MLS",
      "Pool" = "MLS"
    )
  )
```


Checking the changes to make sure they are right.

```{r}
dip_changed |> 
  count(club_short, club_short_clean)
```

Now we'll clean up the columns.

```{r}
dip_cleaned <- dip_changed |> 
  select(!club_short) |> 
  rename(club_short = club_short_clean) |> 
  relocate(year, club_short)

dip_cleaned |> head()  
```

## Cleaning Tabula conversions

Each of these files from years 2018 to 2021 will have different issues.

I'm adding the year col as a number

### 2018 salaries

For this one, the dollar signs came in as their own column, which was actually helpful so the numbers were actually numbers. I also need to add the year.

```{r}
#| label: import-2018

s2018_imported <- read_csv(
  "data-raw/tabula-2018-09-15-Salary-Information-Alphabetical.csv",
  col_names = c(
    "club_long",
    "last_name",
    "first_name",
    "position",
    "currency_1",
    "base_salary",
    "currency_2",
    "guaranteed_compensation"
  )
) |> 
  select(!starts_with("currency")) |> # removes dollar signs
  add_column(year = "2018", .before = "club_long")

s2018_imported
```

### 2019 salaries

For this one, our salaries come in with `$` in the salary fields, and also with a trailing space and `)`. We import those a "raw" versions and clean them before using `type_convert()` to have dplyr try to type them again, changing them to numberic. For some reason I couldn't just convert to `as.numeric()` even after it was cleaned up as they came up NA. I also add the year.

```{r}
#| label: import-2019
s2019_imported <- read_csv(
  "data-raw/tabula-2019-Salary-List-Fall-Release-FINAL-Salary-List-Fall-Release-MLS.csv",
  col_names = c(
    "club_long",
    "last_name",
    "first_name",
    "position",
    "base_salary_raw",
    "guaranteed_compensation_raw"
  )
) |> 
  mutate(
    base_salary = str_remove_all(base_salary_raw, "[\\)\\$]") |> str_trim(),
    guaranteed_compensation = str_remove_all(guaranteed_compensation_raw, "[\\)\\$]") |> str_trim()
  ) |> 
  type_convert() |> 
  select(!ends_with("raw")) |> 
  add_column(year = "2019", .before = "club_long")


s2019_imported |> head()
```

### 2020 salaries

```{r}
s2020_imported <- read_csv(
  "data-raw/tabula-2020-Fall-Winter-Salary-List-alphabetical.csv",
  col_names = c(
    "last_name",
    "first_name",
    "club_long",
    "position",
    "base_salary_raw",
    "guaranteed_compensation_raw"
  )
) |> 
  mutate(
    base_salary = str_remove(base_salary_raw, "\\$"),
    guaranteed_compensation = str_remove(guaranteed_compensation_raw, "\\$"),
  ) |> 
  select(!ends_with("raw")) |> 
  type_convert() |> 
  add_column(year = "2020", .before = "last_name")

s2020_imported |> head()
```

### 2021 salaries

```{r}
s2021_imported <- read_csv(
  "data-raw/tabula-2021-MLSPA-Fall-Salary-release (1).csv",
  col_names = c(
    "club_long",
    "last_name",
    "first_name",
    "position",
    "base_salary_raw",
    "guaranteed_compensation_raw"
  )
) |> 
  mutate(
    base_salary = str_remove(base_salary_raw, "\\$"),
    guaranteed_compensation = str_remove(guaranteed_compensation_raw, "\\$"),
  ) |> 
  select(!ends_with("raw")) |> 
  type_convert() |> 
  add_column(year = "2021", .before = "club_long")

s2021_imported |> head()
```

## Import excel data

### 2022 data

```{r}
s2022_imported <- read_excel("data-raw/MLS Salary 2022.xlsx") |> clean_names() |> 
  add_column(year = "2022", .before = "first_name") |> 
  rename(position = position_s,
         club_long = club)

s2022_imported |> glimpse()
```

### 2023 data

```{r}
s2023_imported <- read_excel("data-raw/MLS Salary 2023.xlsx") |> clean_names() |> 
  add_column(year = "2023", .before = "first_name") |> 
  rename(position = position_s,
         club_long = club)

s2023_imported |> glimpse()
```

## Merge non-dip data

The tabula and excel data both use long club names, so I'll put those together.

```{r}
non_dip_data <- s2018_imported |> 
  bind_rows(s2019_imported, s2020_imported, s2021_imported, s2022_imported, s2023_imported)

non_dip_data |> glimpse()
```

## Cleaning non-dip data

Checking the names for errors.

```{r}
non_dip_data |> 
  count(club_long)
```

Some things of note:

- 2021 New England Revolution may be missing last `n`.
- Montreal should probably be CF Montreal

Let's make the changes

```{r}
non_dip_changed <- non_dip_data |> 
  mutate(
    club_long_changed = recode(
      club_long,
      "Montreal" = "CF Montreal",
      "New England Revolutio" = "New England Revolution",
      "DC United" = "D.C. United",
      "Vancouver Whitecaps" = "Vancouver Whitecaps FC",
      "Minnesota United" = "Minnesota United FC"
    )
  )

non_dip_changed  |> 
  count(club_long_changed, club_long)
```

Now we'll fix up the columns.

```{r}
non_dip_cleaned <- non_dip_changed |> 
  select(-club_long) |> 
  rename(club_long = club_long_changed) |> 
  relocate(year, club_long)

non_dip_cleaned |> glimpse()
```

## Building both names

I'd like to have `club_long` and `club_short` for each row of data, but to do that I need to build a conversion chart of some kind. My plan is this:

- Get a list of all the short names
- Get a list of all the long names
- Add them both to a Google Sheet and match them manually
- Download that sheet and then use joins to fill gaps

I used the following to get the names, though just to copy into a spreadsheet for a lot of manual research and work.

```{r}
clubs_short <- dip_cleaned |> distinct(year, club_short) |> 
  arrange(year, club_short)

# clubs_short |> clipr::write_clip()
```


Let's get the long names:

```{r}
clubs_long <- non_dip_data |> distinct(year, club_long) |> 
  arrange(year, club_long)

# clubs_long |> clipr::write_clip()
```

### Pulling in the spreadsheet

Download the clubs list. Commented out after saved locally.

```{r}
# download.file("https://docs.google.com/spreadsheets/d/e/2PACX-1vQqXJxbbrBsikirZrGyXYV_G6cFZp_dYmcf52UfSYM7Kw3akGlkO5jKP8ZL8WtRA5qUJgFMNPG8JYov/pub?output=csv", "data-processed/clubs-list.csv")

clubs_list <- read_csv("data-processed/clubs-list.csv") |> 
  mutate(year = as.character(year))

clubs_list |> 
  count(club_short, club_long)
```

#### Fill in our dip

dip data doesn't have club_long. We'll join with our master clubs list to get them.

```{r}
dip_full <- dip_cleaned |> 
  left_join(clubs_list, join_by(year == year, club_short == club_short))

dip_full |> 
  filter(is.na(club_long))
```

#### Fill in non-dip

```{r}
non_dip_full <- non_dip_cleaned |> 
  left_join(clubs_list, join_by(year == year, club_long == club_long))

non_dip_full |> 
  filter(is.na(club_short))
```

## Merge it all together

```{r}
mls_salaries <- dip_full |> 
  bind_rows(non_dip_full) |> 
  select(
    year,
    club_short,
    last_name,
    first_name,
    position,
    base_salary,
    guaranteed_compensation,
    club_long,
    conference
  )

mls_salaries |> glimpse()
```

## Export the cleaned data