---
title: "Analysis 2023"
---

## Goals of this notebook

We'll explore MLS Salaries through history. We'll start with the most recent data, then look back historically. A couple of questions that come to mind:

- Which players are getting paid the most this year?
- Which teams have the highest salary bill this year?
- How do team salary rankings compare over time?

[Per the MLS Player's Association](https://mlsplayers.org/resources/salary-guide), "compensation" is: The Annual Average Guaranteed Compensation (Guaranteed Comp) number includes a player's base salary and all signing and guaranteed bonuses annualized over the term of the player's contract, including option years.

## Setup

```{r}
#| label: setup
#| echo: true
#| results: hide
#| message: false
#| warning: false

library(tidyverse)
library(janitor)
library(scales)
library(ggrepel)
library(DT)
```


## Import

Getting the cleaned data.

```{r}
salaries <- read_rds("data-processed/mls-salaries.rds")

salaries |> glimpse()
```

### Setting the most recent year of data

I'm creating an object called `salaries_new` because at some point I'll have new data and might want to just change the year.

```{r}
recent_year <- "2023"
```


## Players with highest salaries

For all players, all time. This is based on compensation. Printing search able table.

```{r}
sal_high <- salaries |> 
  arrange(compensation |> desc()) |> 
  select(!c(club_long, conference, base_salary))

# top of list
sal_high |> 
  filter(compensation >= 7000000)
```

For the most recent year. Again just the top of the list.

```{r}
sal_high |> 
  filter(year == recent_year) |> 
  filter(compensation >= 4000000)
```

For 2023, it looks like Austin FC's Sebastián Driussi is the 6th highest payer, and Houston's Hector Herrera is the 7th highest-paid player.

### Difference with just base pay in 2023?

```{r}
sal_high_base <- salaries |> 
  arrange(base_salary |> desc()) |> 
  select(!c(club_long, conference, compensation))

sal_high_base |> filter(year == recent_year, base_salary >= 2000000) 
```


## Team salaries

We'll get per-year salaries by team, then look at just this year.

```{r}
sal_team <- salaries |> 
  group_by(year, club_long) |> 
  summarise(total_compensation = sum(compensation)) |> 
  arrange(total_compensation |> desc())

# peek at the top
sal_team |> head(10)
```

And let's look at this year

```{r}
sal_team_recent <- sal_team |> filter(year == recent_year)

# peek
sal_team_recent |> head(10)
```

Let's round the numbers for our chart.

```{r}
sal_team_recent_mil <- sal_team_recent |> 
  mutate(total_millions = (total_compensation / 1000000) |> round(1))

sal_team_recent_mil
```



Let's chart this

```{r}
#| label: fig-2023-team-salary
#| fig-height: 8
#| fig-cap: "MLS team salaries"
#| fig-alt: "Bar chart showing MLS team salaries from highest to lowest. Toronto FC tops the list at XXX."


sal_team_recent_mil |> 
  filter(club_long != "Major League Soccer") |> 
  ggplot(mapping = aes(
    x = total_millions,
    y = club_long |> reorder(total_compensation)
  )) +
  geom_bar(stat='identity') +
  geom_text(aes(label = paste("$", as.character(total_millions), sep = "")), color = "white", hjust = 1.25) +
  scale_x_continuous(labels = label_dollar()) +
  labs(
    x = "Total team spending in $ millions",
    y = "",
    title = "2023 MLS Team Salaries",
    subtitle = str_wrap("Salaries includes each player's base salary\
                        plus all signing and guaranteed bonuses annualized\
                        over the term of the player's contract, including option years.")
  ) +
  theme_minimal()
```

## Team spending over time

Let's look at team spending over the past five years. To do this, we have to create a ranking for the spending.

- I'm removing players not affiliated with teams
- When I added a third column to the group because I wanted to use long names for something, the ranking broke. I had to break the group then use the `.by` argument for `rank()`.


```{r}
sal_team_rank <- salaries |> 
  filter(club_short != "MLS" | club_short |> is.na()) |> 
  group_by(year, club_short, club_long) |> 
  summarise(
    total_comp = sum(compensation, na.rm = TRUE)
  ) |> 
  arrange(year, total_comp |> desc()) |> 
  ungroup() |> #<1>
  mutate(rank = rank(-total_comp), .by = year) #<2>

# peek
sal_team_rank |> head(20)
```


1. I break the `group_by` here.
2. Then I set the ranking to work by year.

Visualizing all of them would be tricky. Let's do the top five over last five years.


```{r}
sal_team_rank_top <- sal_team_rank |> 
  filter(rank <= 5,
         year >= "2019")

sal_team_rank_top
```

Peek at this a different way

```{r}
sal_team_rank_top |> 
  select(-total_comp) |> 
  pivot_wider(names_from = year, values_from = rank)
```

### Let's visualze spending rank

I want to use a list of team-specific colors. I'm trying this manually first, based on the results of the chart. The colors were pulled [from here](https://teamcolorcodes.com/soccer/mls-team-color-codes/), though I wish I could instead join with a data package and pull from the team colors.

```{r}
#| label: club_colors

club_color_list <- c(
  "#80000A", # ATL
  "#00B140", # AUS
  "#7CCDEF", # CHI
  "#FEDD00", # CLB
  "#EF3E42", # DC
  "#00245D", # LA
  "#C39E6D", # LAFC
  "#F7B5CD", # MIA
  "#CE0E2D", # NE
  "#5D9741", # SEA
  "#B81137"  # TFC
)
```


And now the chart

```{r}
#| label: chart_rank_t5

sal_team_rank_top |> 
  # filter()
  ggplot(aes(x=year, y=rank, group = club_short)) +
  geom_line(aes(color = club_short)) +
  geom_point(aes(color = club_short, size = 3)) +
  geom_label_repel(aes(label = club_short), size = 3) +
  scale_y_reverse() +
  scale_colour_manual(values = club_color_list) +
  scale_size_continuous(guide = "none") +
  labs(
    title = "MLS Spending Rank",
    subtitle = "The top five spending clubs over the past five years, by rank.",
    color = "Club",
    x = NULL,
    y = "Spending Rank"
  ) +
  theme_minimal()
```


