---
title: "Points vs salaries"
---

## Goal of this notebook

Does spending more on players ensure a better season? I want to compare season points to salaries spent by MLS teams.

I thought I had found some [standings data on kaggle](https://www.kaggle.com/datasets/josephvm/major-league-soccer-dataset?resource=download&select=all_tables.csv), but it only goes through part of 2021.

I worked it out using 2020 data, but there is nothing to really show.

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
options(dplyr.summarise.inform = FALSE)
```

## Importing

```{r}
#| label: import
#| message: false

mls_standings <- read_csv("data-raw/all_tables.csv") |> clean_names()
mls_salaries <- read_rds("data-processed/mls-salaries.rds")
```


## Salaries

Starting with salaries data since this is "cleaned" already.

```{r}
sal_20 <- mls_salaries |> 
  filter(year == "2020") |> 
  group_by(club_long, conference) |> 
  summarise(team_salary = sum(compensation))

sal_20
```

## Standings

Checking what we have for standings data.

```{r}
mls_standings |> 
  summarise(
    yr_old = min(year),
    yr_new = max(year)
  )
```

## Trying this with 2020 anyway

### Peek at the data

```{r}
std_20 <- mls_standings |> filter(year == "2020")
```

### Cleaning names

There are some issues here with team names. There are notes on the names and some names are different than what is used with MLS Salaries.



```{r}
std_20_clean <- mls_standings |> 
  filter(year == "2020", conference == "Overall") |> 
  mutate(
    team_clean = str_remove_all(team, regex("\\[.\\]|\\(.\\)")) |> str_trim(),
    team_new = case_match(
      team_clean,
      "Columbus Crew SC" ~ "Columbus Crew",
      "Houston Dynamo" ~ "Houston Dynamo FC",
      "Los Angeles FC" ~ "LAFC",
      "Vancouver Whitecaps FC" ~ "Vancouver Whitecaps",
      .default = team_clean
    )
  ) |>
  arrange(team_new) |> 
  distinct(team_new, pts)


std_20_clean
```



## Put them together

```{r}
combined_20 <- sal_20 |> 
  left_join(std_20_clean, by = join_by(club_long == team_new)) |> 
  drop_na()

combined_20
```

## Plot the relationship

```{r}
ggplot(combined_20) +
  aes(x = pts, y = team_salary) +
  geom_point(aes(color = conference), size = 4) +
  # geom_text(aes(label = club_long)) +
  ggrepel::geom_text_repel(aes(label = club_long), size = 3)
```

There isn't one, really. At least not for this year. It's also difficult to read.







